<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{vignette 1: old three argument syntax}
-->

# Old three argument syntax

This vignette covers the old three argument syntax used in the
`*_named` functions; see the other vignette "recommended variable
argument syntax" for documentation about the `*_variable` functions.

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `str_match_named` and `str_match_all_named` functions take exactly three arguments as input: 
- `subject` is the character vector from which we want to extract
  tabular data.
- `pattern` is the (character scalar) regular expression with named
  capture groups used for extraction. The named capture groups should be literally specified, e.g. `(?P<group_name>subpattern)`
- `fun.list` is a list with names that correspond to capture groups,
  and values are functions used to convert the extracted character
  data to other (typically numeric) types.

## Genomic range example

For example, consider the character vector of subject strings below:

```{r}
subject.vec <- c(
  "chr10:213,054,000-213,055,000",
  "chrM:111,000",
  "this will not match",
  NA, # neither will this.
  "chr1:110-111 chr2:220-222") # two possible matches.
```

With these genomic range subjects, the goal is to extract the chromosome name (before the colon), along with the numeric start and end locations (separated by a dash). That pattern is coded below,

```{r}
chr.pos.pattern <- paste0(
  "(?P<chrom>chr.*?)",
  ":",
  "(?P<chromStart>[0-9,]+)",
  "(?:",
    "-",
    "(?P<chromEnd>[0-9,]+)",
  ")?")
```

Note that it is often preferable (as above) to code the pattern using paste0, in order to put different parts of the pattern on each line. In particular, I recommend putting each named capture group on its own line -- this is one of the ideas that is used in the variable argument syntax (see other vignette for more info).

## Extract first match from each subject

Using the pattern on the subjects above results in

```{r}
(match.mat <- namedCapture::str_match_named(subject.vec, chr.pos.pattern))
str(match.mat)
```

Note that the third argument (list of conversion functions) is omitted in the code above. In that case, the return value is a character matrix, in which missing values indicate missing subjects or no match. The empty string is used for optional groups which are not used in the match (e.g. chromEnd group/column for second subject).

## Third argument: list of type conversion functions

However we often want to extract numeric data -- in this case we want to convert chromStart/End to integers. You can do that by supplying a named list of conversion functions as the third argument. Each function should take exactly one  argument, a character vector (data in the matched column/group), and return a vector of the same size. The code below specifies the `keep.digits` function for both `chromStart` and `chromEnd`.

```{r}
keep.digits <- function(x)as.integer(gsub("[^0-9]", "", x))
conversion.list <- list(chromStart=keep.digits, chromEnd=keep.digits)
(match.df <- namedCapture::str_match_named(
  subject.vec, chr.pos.pattern, conversion.list))
str(match.df)
```

Note that a `data.frame` is returned when the third argument is specified, in order to handle non-character data types returned by the conversion functions.

## Extract all matches from each subject

Note in the examples above that the last subject has two possible
matches, but only the first is returned by `str_match_named`. Use
`str_match_all_named` to get ALL matches in each subject (not just the
first match).

```{r}
namedCapture::str_match_all_named(
  subject.vec, chr.pos.pattern, conversion.list)
```

As shown above, the result is a list with one element for
each subject. Each list element is a data.frame with one row for each
match.

## Named output

If the pattern specifies the `name` group, then it will be used for
the rownames of the output, and it will not be included as a column. For example the pattern below uses `name` for the first group:

```{r}
name.pattern <- paste0(
  "(?P<name>chr.*?)",
  ":",
  "(?P<chromStart>[0-9,]+)",
  "(?:",
    "-",
    "(?P<chromEnd>[0-9,]+)",
  ")?")
try(named.mat <- namedCapture::str_match_named(
  subject.vec, name.pattern, conversion.list))
(named.mat <- namedCapture::str_match_named(
  subject.vec[-(3:4)], name.pattern, conversion.list))
(named.list <- namedCapture::str_match_all_named(
  subject.vec, name.pattern, conversion.list))
```

Note in the above code we use `try` because it is an error if any
`name` groups are missing (and they are for the subjects 3 and 4).

The named output feature makes it easy to select particular elements of the extracted data by name, e.g.

```{r}
named.mat["chr1", "chromStart"]
named.list[[5]]["chr2", "chromStart"]
```

Note that if the subject is named, its names will be used to name the output (rownames or list names).

```{r}
named.subject.vec <- c(
  ten="chr10:213,054,000-213,055,000",
  M="chrM:111,000",
  nomatch="this will not match",
  missing=NA, # neither will this.
  two="chr1:110-111 chr2:220-222") # two possible matches.
namedCapture::str_match_named(
  named.subject.vec, chr.pos.pattern, conversion.list)
namedCapture::str_match_all_named(
  named.subject.vec, chr.pos.pattern, conversion.list)
```

If the subject has names, and the `name` group is specified, then the subject names are used to name the output (and the `name` column is included in the output).

```{r}
named.subject.vec <- c(
  ten="chr10:213,054,000-213,055,000",
  M="chrM:111,000",
  nomatch="this will not match",
  missing=NA, # neither will this.
  two="chr1:110-111 chr2:220-222") # two possible matches.
namedCapture::str_match_named(
  named.subject.vec, name.pattern, conversion.list)
namedCapture::str_match_all_named(
  named.subject.vec, name.pattern, conversion.list)
```

Next, read the "recommended variable argument syntax" vignette for
information about how to use the `*_variable` functions.
