<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{vignette 4: comparing regex functions for data frames}
-->

# Comparing regex functions for data frames

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette I show comparisons between
`namedCapture::df_match_variable` and its closest cousin in the R
package universe, `tidyr::extract`. The two packages can be used to
compute the same result, but the code/syntax is different.

## Longer more readable syntax 

In this first comparison we use a syntax with each group name on the
same line as its pattern. Here are some observations from the
comparison:

* The namedCapture code is shorter. The tidyr code is longer
  mostly because the for loop that you see below for tidyr is hidden
  inside the definition of `namedCapture::df_match_variable`.
* Converting extracted character groups to numeric column types is
  specified via the `convert` argument of `tidyr::extract`, which uses
  `utils::type.convert`. Because `type.convert` does not know how to
  convert strings like `111,000` to integer, we first need to use
  `remove.commas` to create a new data.frame to use as input to
  `tidyr::extract`. In contrast namedCapture supports arbitrary
  group-specific type conversion functions; we specify `to.int` on the
  same line as the corresponding name/pattern for the
  chromStart/chromEnd groups.

```{r}
## First define data.
(sacct.df <- data.frame(
  position=c(
    "chr10:213,054,000-213,055,000",
    "chrM:111,000-222,000",
    "this will not match",
    NA, # neither will this.
    "chr1:110-111 chr2:220-222"), # two possible matches.
  JobID=c(
    "13937810_25",
    "13937810_25.batch",
    "13937810_25.extern",
    "14022192_[1-3]",
    "14022204_[4]"),
  stringsAsFactors=FALSE))
remove.commas <- function(x)gsub(",", "", x)
long.list <- list()

## namedCapture: 29 lines of code.
range.list <- list(
  "\\[",
  task1="[0-9]+", as.integer,
  "(?:-",#begin optional end of range.
  taskN="[0-9]+", as.integer,
  ")?", #end is optional.
  "\\]")
task.list <- list(
  "(?:",#begin alternate
  task="[0-9]+", as.integer,
  "|",#either one task(above) or range(below)
  range.list,
  ")")#end alternate
to.int <- function(x)as.integer(remove.commas(x))
(long.list$namedCapture <- namedCapture::df_match_variable(
  sacct.df,
  JobID=list(
    job="[0-9]+", as.integer,
    "_",
    task.list,
    "(?:[.]",
    type=".*",
    ")?"),
  position=list(
    chrom="chr.*?",
    ":",
    chromStart=".*?", to.int,
    "-",
    chromEnd="[0-9,]*", to.int)))

## tidyr: 46 lines of code.
range.vec <- c(
  "\\[",
  task1="[0-9]+", 
  "(?:-",#begin optional end of range.
  taskN="[0-9]+", 
  ")?", #end is optional.
  "\\]")
task.vec <- c(
  "(?:",#begin alternate
  task="[0-9]+", 
  "|",#either one task(above) or range(below)
  range.vec,
  ")")#end alternate
regex.list <- list(
  JobID=c(
    job="[0-9]+", 
    "_",
    task.vec,
    "(?:[.]",
    type=".*",
    ")?"),
  position=c(
    chrom="chr.*?",
    ":",
    chromStart=".*?",
    "-",
    chromEnd="[0-9,]*"))
tidyr.input <- transform(
  sacct.df,
  position=remove.commas(position))
tidyr.df.list <- list(sacct.df)
for(col.name in names(regex.list)){
  regex.vec <- regex.list[[col.name]]
  is.group <- names(regex.vec)!=""
  format.vec <- ifelse(is.group, "(%s)", "%s")
  group.vec <- sprintf(format.vec, regex.vec)
  regex <- paste(group.vec, collapse="")
  group.names <- names(regex.vec)[is.group]
  result <- tidyr::extract(
    tidyr.input, col.name, group.names, regex, convert=TRUE)
  to.save <- result[, group.names, drop=FALSE]
  names(to.save) <- paste0(col.name, ".", group.names)
  tidyr.df.list[[col.name]] <- to.save
}
names(tidyr.df.list) <- NULL
long.list$tidyr <- do.call(cbind, tidyr.df.list)

## Make sure the results are the same.
t(sapply(long.list, names))
t(sapply(long.list, sapply, class))
long.list$tidyr$JobID.type <- ifelse(
  is.na(long.list$tidyr$JobID.type),
  "",
  long.list$tidyr$JobID.type)
with(long.list, identical(tidyr, namedCapture))

```

Exercise for the reader use `rematch2::bind_re_match` instead of
`tidyr::extract` (you should only have to change a few lines of code
in the for loop).

## Shorter less readable syntax

This second comparison uses a syntax with the entire regex on one
line. In my opinion this syntax makes the regular expressions more
difficult to read/understand. Complicated regular expressions like the
one used for matching the JobID column are not
maintainable/understandable at all using this syntax.

```{r}
## First define data.
(sacct.df <- data.frame(
  position=c(
    "chr10:213,054,000-213,055,000",
    "chrM:111,000-222,000",
    "this will not match",
    NA, # neither will this.
    "chr1:110-111 chr2:220-222"), # two possible matches.
  JobID=c(
    "13937810_25",
    "13937810_25.batch",
    "13937810_25.extern",
    "14022192_[1-3]",
    "14022204_[4]"),
  stringsAsFactors=FALSE))
short.list <- list()

## tidyr alternate (13 lines total)
e <- function(col.name, group.names, pattern){
  result <- tidyr::extract(
    sacct.df, col.name, group.names, pattern, convert=TRUE)
  to.save <- result[, group.names, drop=FALSE]
  names(to.save) <- paste0(col.name, ".", group.names)
  to.save
}
short.list$tidyr <- do.call(cbind, list(
  sacct.df,
  e("JobID", c("job", "task", "task1", "taskN", "type"),
    "([0-9]+)_(?:([0-9]+)|\\[([0-9]+)(?:-([0-9]+))?\\])(?:[.](.*))?"),
  e("position", c("chrom", "chromStart", "chromEnd"),
    "(chr.*?):(.*?)-([0-9,]*)")))

## namedCapture alternate (7 lines total)
(short.list$namedCapture <- namedCapture::df_match_variable(
  sacct.df,
  JobID="(?P<job>[0-9]+)_(?:(?P<task>[0-9]+)|\\[(?P<task1>[0-9]+)(?:-(?P<taskN>[0-9]+))?\\])(?:[.](?P<type>.*))?",
  position="(?P<chrom>chr.*?):(?P<chromStart>.*?)-(?P<chromEnd>[0-9,]*)"))
for(N in names(short.list$namedCapture)){
  short.list$namedCapture[[N]] <- type.convert(short.list$namedCapture[[N]], as.is=TRUE)
}

## Make sure the results are the same.
t(sapply(short.list, names))
t(sapply(short.list, sapply, class))
short.list$tidyr$JobID.type <- ifelse(
  is.na(short.list$tidyr$JobID.type),
  "",
  short.list$tidyr$JobID.type)
with(short.list, identical(tidyr, namedCapture))

```

## Comparison with rematch2

`rematch2::bind_re_match` is similar to `tidyr::extract` but
additionally supports named capture regular expressions. Overall the
comparison shows that both packages can use a relatively verbose and
readable syntax to define complex regular expressions piece by piece:

```{r}
range.list <- list(
  "\\[",
  task1="[0-9]+", as.integer,
  list(
    "-",#begin optional end of range.
    taskN="[0-9]+", as.integer
  ), "?", #end is optional.
  "\\]")
namedCapture::df_match_variable(sacct.df, JobID=range.list)

range.pat <- paste0(
  "\\[",
  "(?<task1>[0-9]+)", 
  "(?:",
  "-",#begin optional end of range.
  "(?<taskN>[0-9]+)",
  ")?", #end is optional.
  "\\]")
rematch2::bind_re_match(sacct.df, JobID, range.pat)
task.list <- list(
  "_",
  list(
    task="[0-9]+", as.integer,
    "|",#either one task(above) or range(below)
    range.list))
namedCapture::df_match_variable(sacct.df, JobID=task.list)

task.pat <- paste0(
  "_",
  "(?:",
  "(?<task>[0-9]+)", 
  "|", #either one task(above) or range(below)
  range.pat,
  ")")
rematch2::bind_re_match(sacct.df, JobID, task.pat)

job.list <- list(
  job="[0-9]+", as.integer,
  task.list,
  list(
    "[.]",
    type=".*"
  ), "?")
(job.namedCapture <- namedCapture::df_match_variable(sacct.df, JobID=job.list))

job.pat <- paste0(
  "(?<job>[0-9]+)", 
  task.pat,
  "(?:",
  "[.]",
  "(?<type>.*)",
  ")?")
(job.rematch2 <- rematch2::bind_re_match(sacct.df, JobID, job.pat))

pos.namedCapture <- namedCapture::df_match_variable(
  job.namedCapture, position=list(
    chrom="chr.*?",
    ":",
    chromStart=".*?", to.int,
    "-",
    chromEnd="[0-9,]*", to.int))
str(pos.namedCapture)

pos.rematch2 <- rematch2::bind_re_match(
  job.rematch2,  position, paste0(
    "(?<chrom>chr.*?)",
    ":",
    "(?<chromStart>.*?)", 
    "-",
    "(?<chromEnd>[0-9,]*)"))
str(pos.rematch2)

```

The main difference in syntax is that group names are specified in
the regular expression string literal for rematch2, whereas group
names are specified as R argument names for namedCapture
A difference in the result is that all columns of `pos.rematch2` are
character, whereas some columns of `pos.namedCapture` have already
been converted to integer. Using `rematch2` type conversion may be
accomplished as a post-processing step:

```{r}
converted.rematch2 <- transform(
  pos.rematch2,
  JobID.job=to.int(job),
  JobID.task1=to.int(task1),
  JobID.taskN=to.int(taskN),
  JobID.task=to.int(task),
  JobID.type=type,
  position.chrom=chrom,
  position.chromStart=to.int(chromStart),
  position.chromEnd=to.int(chromEnd),
  stringsAsFactors=FALSE)
some.rematch2 <- converted.rematch2[, names(pos.namedCapture)]
identical(some.rematch2, pos.namedCapture)
```

Exercise for the reader: convert all the `rematch2::bind_re_match`
calls in this section to `tidyr::extract` calls.
